Visualizing Contingency Tables

### 2006-12-07 trunk/vcd/paper/Shadings/JCGS.Rnw
incorporated changes for second revision of JCGS shadings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@624 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Packages:
\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}
%% need no \usepackage{Sweave}

@@

myseed <- 1071
nrep <- 5000
nrep_char <- "5,000"
recompute <- TRUE
@

@@

  set.seed(myseed)
  pun_dmax <- coindep_test(pun, 3:4, n = nrep)
  set.seed(myseed)
  pun_maxchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  pun_sumchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
  set.seed(myseed)
  pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
    varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)

  save(art_max, art_chisq,
       pring_max, pring_chisq,
       alz_dmax, alz_maxchisq, alz_sumchisq, alz_cotab,
       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab,
       file = "JCGS-inference.rda")
} else {
  load("JCGS-inference.rda")
}
@

@@

However, the sum of squares is not the only plausible way of capturing deviations
from zero in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the `dependence' (i.e., significant
departure from independence), if any, is the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute values exceed $c_\alpha$ violate the null hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

@@

\setkeys{Gin}{width=2\textwidth}
\begin{figure}[p]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=12,width=16,results=hide,include=FALSE>>=
mymar <- c(1.5, 0.5, 0.5, 2.5)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 3)))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = mymar,
  gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
mosaic(art, newpage = FALSE, gp = shading_hsv, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 3))
mosaic(art, newpage = FALSE, gp = shading_hcl, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()

pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
mosaic(pring, newpage = FALSE, gp = shading_Friendly, margins = mymar,
  gp_args = list(eps = NULL, lty = 1, interpolate = c(1, 1.5)))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
mosaic(pring, newpage = FALSE, gp = shading_hsv, margins = mymar,
  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 3))
mosaic(pring, newpage = FALSE, gp = shading_hcl, margins = mymar,
  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
popViewport(2)
@
\includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
\caption{Upper row: Mosaic plot for the arthritis data with Friendly shading (left),
HSV maximum shading (middle), HCL maximum shading (right). Lower row: Mosaic
plot for the piston rings data with fixed user-defined cut offs 1 and 1.5 and 
Friendly shading (left), HSV sum-of-squares shading (middle), HCL sum-of-squares shading
(right).}
\label{fig:shadings}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children. The contingency table comprises four margins:
memory of punishments as a child (yes/no), attitude as a binary variable
(approval of ``moderate'' punishment or ``no'' approval), highest level
of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).
It is of interest whether there is an association between memories of corporal punishments
as a child and attitude towards punishment of children as an adult, controlling
for age and education. Figure~\ref{fig:pun} shows a conditional association
plot of memory and attitude given age and education. This can be
interpreted as a diagnostic residual plot for the associated log-linear conditional
independence model. It shows an association between memories and attitude
for the lowest education group (first column) and highest age group (last row):
experienced violence seems to engender violence again as there are less adults
that disapprove punishment in the group with memories of punishments than expected
under independence. For the remaining four age-education groups, there seems to be
no association: all residuals of the conditional independence model are very close to
zero in these cells. All three tests agree again that there is significant
departure from conditional independence in this table: the sum-of-squares
statistic is $\sum_k X_k^2 = \Sexpr{stat(pun_sumchisq)}$ ($p = \Sexpr{pval(pun_sumchisq)}$),
the maximum sum-of-squares statistic is $\max_k X_k^2 = \Sexpr{stat(pun_maxchisq)}$
($p = \Sexpr{pval(pun_maxchisq)}$), and the double maximum statistic
is $\max_k M_k = \Sexpr{stat(pun_dmax)}$ ($p = \Sexpr{pval(pun_dmax)}$). The $\max_k X_k^2$
result is visualized by means of a maximum sum-of-squares shading in Figure~\ref{fig:pun}
with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
The full-color palette is used only for those strata associated with a
sum-of-squares statistic $X_k^2$ significant at (overall) 5\% level, the reduced-color
palette is used otherwise. This highlights that the dependence pattern is significant
only for the middle and high age group in the low education column. The other panels
in the first column and last row also show a similar dependence pattern, however, it is not
significant at 5\% level and hence graphically down-weighted by using reduced color.

@@

## Figure 5
set.seed(rseed)
cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
  n = 5000, type = "assoc", test = "maxchisq", interpolate = 1:2)
\end{verbatim}

>>>>

%% Packages:
\usepackage{graphicx,color,a4wide,thumbpdf,rotating,hyperref}
%% need no \usepackage{Sweave}

@@

myseed <- 1071
nrep <- 5000
nrep_char <- "5,000"
recompute <- FALSE
@

@@

  set.seed(myseed)
  pun_dmax <- coindep_test(pun, 3:4, n = nrep)
  set.seed(myseed)
  pun_maxchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  pun_sumchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
  set.seed(myseed)
  pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
    varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)
  set.seed(myseed)
  pun_cotab2 <- cotab_coindep(pun, condvars = 3:4, type = "mosaic", n = nrep,
    varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)

  save(art_max, art_chisq,
       pring_max, pring_chisq,
       alz_dmax, alz_maxchisq, alz_sumchisq, alz_cotab,
       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab, pun_cotab2,
       file = "JCGS-inference.rda")
} else {
  load("JCGS-inference.rda")
}
@

@@

However, the sum of squares is not the only plausible way of capturing deviations
from zero in the residuals. There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$---and without further 
specification of a certain pattern of dependence no functional $\lambda(\cdot)$
uniformly dominates all others in terms of power of the resulting test procedure.
Therefore, the choice of the functional is usually also guided by the data analysis problem at
hand: one functional which is particularly suitable for identifying the cells responsible for
the `dependence' (i.e., significant departure from independence), if any, is the maximum of
the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute values exceed $c_\alpha$ violate the null hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

@@

\setkeys{Gin}{width=2\textwidth}
\begin{sidewaysfigure}[p]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=12,width=18,results=hide,include=FALSE>>=
##                                       width=16
mymar <- c(1.5, 0.5, 0.5, 2.5)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 3)))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = mymar,
  gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
mosaic(art, newpage = FALSE, gp = shading_hsv, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 3))
mosaic(art, newpage = FALSE, gp = shading_hcl, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()

pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
mosaic(pring, newpage = FALSE, gp = shading_Friendly, margins = mymar,
  gp_args = list(eps = NULL, lty = 1, interpolate = c(1, 1.5)))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
mosaic(pring, newpage = FALSE, gp = shading_hsv, margins = mymar,
  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 3))
mosaic(pring, newpage = FALSE, gp = shading_hcl, margins = mymar,
  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
popViewport(2)
@
%%\includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
\includegraphics[width=.9\textwidth,keepaspectratio]{JCGS-shadings}
\caption{Upper row: Mosaic plot for the arthritis data with Friendly shading (left),
HSV maximum shading (middle), HCL maximum shading (right). Lower row: Mosaic
plot for the piston rings data with fixed user-defined cut offs 1 and 1.5 and 
Friendly shading (left), HSV sum-of-squares shading (middle), HCL sum-of-squares shading
(right).}
\label{fig:shadings}
\end{center}
\end{sidewaysfigure}
\setkeys{Gin}{width=\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<conditional_mosaic2,fig=TRUE,echo=FALSE,height=11,width=11>>=
cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab2)
@
\caption{Conditional mosaic plot with maximum sum-of-squares shading for conditional
independence of memory and attitude given age and education.}
\label{fig:pun2}
\end{center}
\end{figure}

As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children. The contingency table comprises four margins:
memory of punishments as a child (yes/no), attitude as a binary variable
(approval of ``moderate'' punishment or ``no'' approval), highest level
of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).
It is of interest whether there is an association between memories of corporal punishments
as a child and attitude towards punishment of children as an adult, controlling
for age and education. Figure~\ref{fig:pun} shows a conditional association
plot of memory and attitude given age and education. This can be
interpreted as a diagnostic residual plot for the associated log-linear conditional
independence model. 
%FIXME: with mosaic%
Alternatively, a conditional mosaic plot as in Figure~\ref{fig:pun2} can be used for
visualizing the contingency table and its associated residuals from the conditional independence
model using the same shading. Both reveal
%FIXME: without mosaic% It shows 
an association between memories and attitude
for the lowest education group (first column) and highest age group (last row):
experienced violence seems to engender violence again as there are less adults
that disapprove punishment in the group with memories of punishments than expected
under independence. For the remaining four age-education groups, there seems to be
no association: all residuals of the conditional independence model are very close to
zero in these cells. All three tests agree again that there is significant
departure from conditional independence in this table: the sum-of-squares
statistic is $\sum_k X_k^2 = \Sexpr{stat(pun_sumchisq)}$ ($p = \Sexpr{pval(pun_sumchisq)}$),
the maximum sum-of-squares statistic is $\max_k X_k^2 = \Sexpr{stat(pun_maxchisq)}$
($p = \Sexpr{pval(pun_maxchisq)}$), and the double maximum statistic
is $\max_k M_k = \Sexpr{stat(pun_dmax)}$ ($p = \Sexpr{pval(pun_dmax)}$). The $\max_k X_k^2$
result is visualized by means of a maximum sum-of-squares shading in Figure~\ref{fig:pun}
with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
The full-color palette is used only for those strata associated with a
sum-of-squares statistic $X_k^2$ significant at (overall) 5\% level, the reduced-color
palette is used otherwise. This highlights that the dependence pattern is significant
only for the middle and high age group in the low education column. The other panels
in the first column and last row also show a similar dependence pattern, however, it is not
significant at 5\% level and hence graphically down-weighted by using reduced color.

@@

## Figure 5
set.seed(rseed)
cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
  n = 5000, type = "assoc", test = "maxchisq", interpolate = 1:2)

## Figure 6
set.seed(rseed)
cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
  n = 5000, type = "mosaic", test = "maxchisq", interpolate = 1:2)
\end{verbatim}
